Automatically Splitting a Two-Stage Lambda Calculus

### 2015-07-10 paper/abstract.tex
Make some notes on missing points.

Staged programming languages assign a stage to each expression and
evaluate each expression in its assigned stage,
and can be used to describe programs where inputs arrive at different times or rates.

In this paper we present an 
algorithm for statically {\em splitting} these mixed-staged programs 
into two dependent, unstaged programs where the outputs of the first part can be efficiently reused accorss multiple invocations of the second program.

While previous algorithms for performing this transformation 
(also called \emph{pass separation} and \emph{data specialization}) operate on simpler imperative languages,
out algorithm operates on an explicitly-two-staged, typed lambda calculus \lang, 
with a $\fut$ modality denoting computation at a later stage,
and a $\curr$ modality noting purely first stage code.
Notably, our algorithm is the first to with support for mixed-stage recursive and higher-order functions.

We have proven the dynamic correctness of our splitting algorithm with respect
to a partial-evaluation semantics, and mechanized this proof in Twelf.

We have implemented the algorithm in a prototype compiler,
and we show that the ability to split a language with recursion and higher-order features
can enabled non-trivial algorithmic transformations that asymptotically improve code efficiency by nearly a linear factor,
as well as modular expression of staged programming,

>>>>

Staged programming languages assign a stage to each program expression
and evaluate each expression in its assigned stage. A common use of
staged languages is to describe programs where inputs arrive at
different times or rates. In this paper we present an algorithm for
statically {\em splitting} these mixed-staged programs into two
unstaged, but dependent, programs where the outputs of the first
program can be efficiently reused across multiple invocations of the
second. While previous algorithms for performing this transformation
(also called \emph{pass separation} and \emph{data specialization})
were limited to operate on simpler, imperative languages, we define a
splitting algorithm for an explicitly-two-stage, typed lambda
calculus \lang\ with a $\fut$ modality denoting computation at a later
stage, and a $\curr$ modality noting purely first stage code.  Most
notably, the algorithm splits mixed-stage recursive and higher-order
functions.  We have proven the dynamic correctness of our splitting
algorithm with respect to a partial-evaluation semantics, and
mechanized this proof in Twelf. We have also implemented the algorithm
in a prototype compiler, and demonstrate that the ability to split
programs in a language featuring recursion and higher-order features
enables non-trivial algorithmic transformations that asymptotically
improve code efficiency by nearly a linear factor, and also
facilitates modular expression of staged programs.

% todo: static transformation carried out prior to any of the programs
% inputs being available, and thus has low runtime overhead (and,
% unline partial evaluation, is applicable when dynamic code
% generation is not possible)
